---
title: "lab-10: Distances and the Border Zone"
subtitle: 'Ecosystem Science & Sustainability 330'
author: 
 - name: Megan Hoover
format: html
---

Load Libraries:
```{r}
# spatial data science
library(tidyverse)
library(sf)
library(units)

# Data
library(AOI)

# Visualization
library(gghighlight)
library(ggrepel)
library(knitr)
library(flextable)
```


# 1.1 Define a Projection

For this lab we want to calculate distances between features, therefore we need a projection that preserves distance at the scale of CONUS. For this, we will use the North America Equidistant Conic:
```{r}
eqdc <- '+proj=eqdc +lat_0=40 +lon_0=-96 +lat_1=20 +lat_2=60 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs'

#This PROJ.4 string defines an Equidistant Conic projection with the following parameters:

#+proj=eqdc → Equidistant Conic projection
#+lat_0=40 → Latitude of the projection’s center (40°N)
#+lon_0=-96 → Central meridian (96°W)
#+lat_1=20 → First standard parallel (20°N)
#+lat_2=60 → Second standard parallel (60°N)
#+x_0=0 → False easting (0 meters)
#+y_0=0 → False northing (0 meters)
#+datum=NAD83 → Uses the North American Datum 1983 (NAD83)
#+units=m → Units are in meters
#+no_defs → No additional default parameters from PROJ’s database

```


This projection is commonly used for mapping large areas with an east-west extent, especially in North America, as it balances distortion well between the two standard parallels.

# 1.2 Get USA state boundaries

In R, USA boundaries are stored in the AOI package. In case this package and data are not installed:
```{r}
remotes::install_github("mikejohnson51/AOI")
```
Once installed:

USA state boundaries can be accessed with aoi_get(state = 'conus').
Make sure the data is in a projected coordinate system suitable for distance measurements at the national scale (eqdc).
```{r}
#sf objects with state boundaries
states_conus <- aoi_get(state = "conus")

#check current coordinate system: geographic coordinates in degrees, not suitable for measuring distances.
st_crs(states_conus) #returns WGS 84, coords in degrees

#edqc projection is suitable for distance calculations across CONUS

#Now transform the spatial object to use the EQDC projection:
states_eqdc <- st_transform(states_conus, crs = eqdc)

#check cureent coordinate system now: units should be in m, good for distance calculations
st_crs(states_eqdc)


```


# 1.3 Get country boundaries for Mexico, the United States of America, and Canada

In R, country boundaries are stored in the AOI package.

Country boundaries can be accessed with aoi_get(country = c("MX", "CA", "USA")).
Make sure the data is in a projected coordinate system suitable for distance measurements at the national scale (eqdc).
```{r}
# get country boundaries
country_boundaries<- aoi_get(country = c("MX", "CA", "USA"))

#check structure
print(country_boundaries)

#check coordinate system
st_crs(country_boundaries) # (WGS 84) not ideal for distance.

#transforms your spatial object to the EQDC projection.
countries_eqdc <- st_transform(country_boundaries, crs = eqdc)


```


# 1.4 Get city locations from the CSV file

First go to this site and download the appropriate (free) dataset into the data directory of this project.

Once downloaded, read it into your working session using readr::read_csv() and explore the dataset until you are comfortable with the information it contains.

While this data has everything we want, it is not yet spatial. Convert the data.frame to a spatial object using st_as_sf and prescribing the coordinate variables and CRS (Hint what projection are the raw coordinates in?)

Finally, remove cities in states not wanted and make sure the data is in a projected coordinate system suitable for distance measurements at the national scale:

Congratulations! You now have three real-world, large datasets ready for analysis.
```{r}
# Read the CSV into R
cities_df <- read_csv("C:/Users/samho/Desktop/ESS330/github/csu-ess-lab10/data/uscities.csv")

# View structure and columns
glimpse(cities_df)


#Use sf::st_as_sf() to convert the data frame to an sf object using the latitude and longitude columns
# Convert to sf object (WGS 84 = EPSG:4326)
cities_sf <- st_as_sf(cities_df, coords = c("lng", "lat"), crs = 4326)

# Optional: Define CONUS state abbreviations
conus_states <- c(state.abb, "DC")  # 48 states + D.C.

# Filter cities to only CONUS states
cities_conus <- cities_sf[cities_sf$state_id %in% conus_states, ]

# Reproject to EQDC
cities_eqdc <- st_transform(cities_conus, crs = eqdc)

#check crs
st_crs(cities_eqdc)

```


# Question 2:

Here we will focus on calculating the distance of each USA city to (1) the national border (2) the nearest state border (3) the Mexican border and (4) the Canadian border. You will need to manipulate you existing spatial geometries to do this using either st_union or st_combine depending on the situation. In all cases, since we are after distances to borders, we will need to cast (st_cast) our MULTIPPOLYGON geometries to MULTILINESTRING geometries. To perform these distance calculations we will use st_distance().


# 2.1 Distance to USA Border (coastline or national) (km)

For 2.2 we are interested in calculating the distance of each USA city to the USA border (coastline or national border). To do this we need all states to act as single unit. Convert the USA state boundaries to a MULTILINESTRING geometry in which the state boundaries are resolved. Please do this starting with the states object and NOT with a filtered country object. In addition to storing this distance data as part of the cities data.frame, produce a table (flextable) documenting the five cities farthest from a state border. Include only the city name, state, and distance.
```{r}
# Union all state boundaries into a single polygon
usa_union <- st_union(states_eqdc)

# Convert to a MULTILINESTRING for border geometry
usa_border <- st_cast(usa_union, "MULTILINESTRING")

#Calculate distance from each city to national border (in kilometers)
dist_to_border_m <- st_distance(cities_eqdc, usa_border) 
cities_eqdc$dist_to_border_km <- as.numeric(dist_to_border_m) / 1000 #convert m to km

# Create a summary table of the 5 most inland cities
top5_far_cities <- cities_eqdc %>%
  arrange(desc(dist_to_border_km)) %>%
  slice(1:5) %>%
  select(city, state_id, dist_to_border_km)

flextable(top5_far_cities)

```


# 2.2 - Distance to States (km)

For 2.1 we are interested in calculating the distance of each city to the nearest state boundary. To do this we need all states to act as single unit. Convert the USA state boundaries to a MULTILINESTRING geometry in which the state boundaries are preserved (not resolved). In addition to storing this distance data as part of the cities data.frame, produce a table (flextable) documenting the five cities farthest from a state border. Include only the city name, state, and distance.
```{r}
#Use st_combine() instead of st_union() (so borders between states remain),
#Convert the result to MULTILINESTRING to represent all state boundaries,
#Then compute distances from cities to those lines.

# Combine geometries without resolving internal boundaries
state_borders <- states_eqdc %>%
  st_combine() %>%
st_cast("MULTILINESTRING") #represents all state lines, including interior borders.


#Calculate distance from each city to the nearest state boundary
dist_to_state_border_m <- st_distance(cities_eqdc, state_borders)
cities_eqdc$dist_to_state_km <- as.numeric(dist_to_state_border_m) / 1000 #m to km

#Identify the 5 cities farthest from any state boundary
top5_far_from_state <- cities_eqdc %>%
  arrange(desc(dist_to_state_km)) %>%
  slice(1:5) %>%
  select(city, state_id, dist_to_state_km)

flextable(top5_far_from_state)
```


# 2.3 - Distance to Mexico (km)

For 2.3 we are interested in calculating the distance of each city to the Mexican border. To do this we need to isolate Mexico from the country objects. In addition to storing this data as part of the cities data.frame, produce a table (flextable) documenting the five cities farthest from a state border. Include only the city name, state, and distance.
```{r}
# using existing countries_eqdc object, which includes USA, Canada, and Mexico, all projected in EQDC.

# Filter for Mexico only
mexico <- countries_eqdc %>%
  filter(sovereignt == "Mexico")  # use column name sovereignt == "Mexico"

# Convert Mexico to a border line
mexico_border <- mexico %>%
  st_union() %>%
  st_cast("MULTILINESTRING") #MULTILINESTRING to represent the border outline of Mexico

# Calculate distances from cities to Mexico
dist_to_mexico_m <- st_distance(cities_eqdc, mexico_border)
cities_eqdc$dist_to_mexico_km <- as.numeric(dist_to_mexico_m) / 1000 #m to km

# Create a table of the 5 cities farthest from the Mexican border
top5_far_from_mexico <- cities_eqdc %>%
  arrange(desc(dist_to_mexico_km)) %>%
  slice(1:5) %>%
  select(city, state_id, dist_to_mexico_km)

flextable(top5_far_from_mexico)
```


# 2.4 Distance to Canada (km)

For 2.4 we are interested in calculating the distance of each city to the Canadian border. To do this we need to isolate Canada from the country objects. In addition to storing this data as part of the cities data.frame, produce a table (flextable) documenting the five cities farthest from a state border. Include only the city name, state, and distance.
```{r}
# Filter for Canada, Convert Canada polygon to border lines
canada_border <- countries_eqdc %>%
  filter(name == "Canada") %>%
  st_union() %>%
  st_cast("MULTILINESTRING")

#Compute distance from each city to the Canadian border
cities_eqdc$dist_to_canada_km <- st_distance(cities_eqdc, canada_border) %>%
  as.numeric() / 1000 #m to km

# Create a table: 5 cities farthest from the Canadian border
top5_far_from_canada <- cities_eqdc %>%
  arrange(desc(dist_to_canada_km)) %>%
  slice(1:5) %>%
  select(city, state_id, dist_to_canada_km)

flextable(top5_far_from_canada)


```


# Question 3:

In this section we will focus on visualizing the distance data you calculated above. You will be using ggplot to make your maps, ggrepl to label significant features, and gghighlight to emphasize important criteria.


# 3.1 Data
Show the 3 continents, CONUS outline, state boundaries, and 10 largest USA cities (by population) on a single map

Use geom_sf to plot your layers
Use lty to change the line type and size to change line width
Use ggrepel::geom_label_repel to label your cities
```{r}
# Filter the 10 Largest U.S. Cities by Population
# Filter the 10 largest cities by population
top_10_cities <- cities_eqdc %>%
  arrange(desc(population)) %>%
  slice(1:10)

 # Prepare Spatial Layers for Plotting
# Continents: already have countries_eqdc for USA, Canada, and Mexico.
# CONUS Outline: This is the outline of the contiguous United States.
# State Boundaries: The states_eqdc layer used earlier.
# countries_eqdc, states_eqdc, and cities_eqdc

# Add coordinates to your top_10_cities data frame before using it in geom_label_repel():
top_10_cities_coords <- top_10_cities %>%
  mutate(
    coords = st_coordinates(geometry),
    x = coords[, 1],
    y = coords[, 2]
  )

#pass x and y to geom_label_repel():



# Create the plot with all the layers
ggplot() +
  # Plot continents (USA, Canada, Mexico)
  geom_sf(data = countries_eqdc, fill = "lightgray", color = "black", lwd = 0.5) +
  
  # Plot CONUS outline (only USA). lty= line type, lwd= line width
  geom_sf(data = states_eqdc, fill = NA, color = "blue", lty = 2, lwd = 0.8) +
  
  # Plot state boundaries
  geom_sf(data = states_eqdc, fill = NA, color = "black", lty = 1, lwd = 0.5) +
  
  # Plot the cities (top 10 largest cities by population)
  geom_sf(data = top_10_cities, aes(color = city), size = 3) +
  
  # Use ggrepel to label the cities. avoids overlapping, but you can adjust padding with box.padding and point.padding
  geom_label_repel(data = top_10_cities_coords, aes(x= x, y=y, label = city), 
                   box.padding = 0.35, point.padding = 0.5,
                   segment.color = "black", size = 3) +
  
  # Customize the theme
  theme_minimal() +
  
  # Add a title
  ggtitle("Map of 10 Largest U.S. Cities by Population") +
  
  # Adjusting the theme
  theme(
    plot.title = element_text(hjust = 0.5, size = 16),
    legend.position = "none"  # Hide legend
  )
```


# 3.2 City Distance from the Border

Create a map that colors USA cities by their distance from the national border. In addition, re-draw and label the 5 cities that are farthest from the border.
```{r}
#Identify the 5 cities farthest from the border:
farthest_5 <- cities_eqdc %>%
  arrange(desc(dist_to_border_km)) %>%
  slice(1:5) %>%
  mutate(
    coords = st_coordinates(geometry),
    x = coords[, 1],
    y = coords[, 2]
  )

# plot the new map
ggplot() +
  # Plot U.S. states
  geom_sf(data = states_eqdc, fill = "white", color = "black", lwd = 0.3) +
  
  # Plot all cities colored by distance to border
  geom_sf(data = cities_eqdc, aes(color = dist_to_border_km), size = 1, alpha = 0.7) +
  
  # Redraw and highlight the 5 farthest cities
  geom_sf(data = farthest_5, color = "red", size = 2) +
  
  # Label those cities
  geom_label_repel(
    data = farthest_5,
    aes(x = x, y = y, label = city),
    size = 3,
    box.padding = 0.3,
    point.padding = 0.5,
    segment.color = "black"
  ) +
  
  # Scale and theme
  scale_color_viridis_c(name = "Distance to Border (km)") +
  theme_minimal() +
  ggtitle("U.S. Cities Colored by Distance from National Border") +
  theme(plot.title = element_text(hjust = 0.5, size = 16))
```


# 3.3 City Distance from Nearest State

Create a map that colors USA cities by their distance from the nearest state border. In addition, re-draw and label the 5 cities that are farthest from any border.
```{r}
# coloring U.S. cities by their distance to the nearest state border, not the national border. Color all cities by dist_to_state_km

#Identify the 5 cities farthest from any state border:
farthest_5_state <- cities_eqdc %>%
  arrange(desc(dist_to_state_km)) %>%
  slice(1:5) %>%
  mutate(
    coords = st_coordinates(geometry),
    x = coords[, 1],
    y = coords[, 2]
  )

#create map
ggplot() +
  # Plot U.S. state outlines
  geom_sf(data = states_eqdc, fill = "white", color = "black", lwd = 0.3) +
  
  # Plot all cities, colored by distance to state border
  geom_sf(data = cities_eqdc, aes(color = dist_to_state_km), size = 1, alpha = 0.7) +
  
  # Highlight the 5 farthest cities
  geom_sf(data = farthest_5_state, color = "darkred", size = 2) +
  
  # Label the 5 cities
  geom_label_repel(
    data = farthest_5_state,
    aes(x = x, y = y, label = city),
    size = 3,
    box.padding = 0.3,
    point.padding = 0.5,
    segment.color = "black"
  ) +
  
  # Color scale and theme
  scale_color_viridis_c(name = "Distance to State Border (km)") +
  theme_minimal() +
  ggtitle("U.S. Cities Colored by Distance from Nearest State Border") +
  theme(plot.title = element_text(hjust = 0.5, size = 16))
```


# 3.4 Equidistance boundary from Mexico and Canada

Here we provide a little more challenge. Use gghighlight to identify the cities that are equal distance from the Canadian AND Mexican border 
 100 km.

In addition, label the five (5) most populous cites in this zone.

Hint: (create a new variable that finds the absolute difference between the distance to Mexico and the distance to Canada)
```{r}
# Create a new column for the absolute distance difference
cities_eqdc <- cities_eqdc %>%
  mutate(
    dist_diff_km = abs(dist_to_canada_km - dist_to_mexico_km)
  )


# Filter for cities within 100 km difference
equal_dist_cities <- cities_eqdc %>%
  filter(dist_diff_km <= 100)

# Get the 5 most populous cities in that zone
top5_equal_dist <- equal_dist_cities %>%
  arrange(desc(population)) %>%
  slice(1:5) %>%
  mutate(
    coords = st_coordinates(geometry),
    x = coords[,1],
    y = coords[,2]
  )

# Plot with ggplot2, geom_sf, and gghighlight.use gghighlight to emphasize only the cities in the equidistant zone:
ggplot(data = cities_eqdc) +
  # Plot all cities with low alpha
  geom_sf(aes(color = dist_diff_km), size = 1, alpha = 0.5) +
  
  # Highlight only cities within 100 km of equal distance
  gghighlight(dist_diff_km <= 100, use_direct_label = FALSE) +
  
  # State borders for context
  geom_sf(data = states_eqdc, fill = NA, color = "black", lwd = 0.3) +
  
  # Label the 5 most populous cities in the zone
  geom_label_repel(
    data = top5_equal_dist,
    aes(x = x, y = y, label = city),
    size = 3,
    box.padding = 0.3,
    point.padding = 0.5,
    segment.color = "black"
  ) +
  
  scale_color_viridis_c(name = "Distance Diff (km)") +
  ggtitle("Cities ~Equidistant from Canada and Mexico (±100 km)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 15))

# map shows:
# Highlight only cities within ±100 km of equal distance to Mexico and Canada
#Label the 5 most populous of those cities

```

