---
title: "lab-10: Distances and the Border Zone"
subtitle: 'Ecosystem Science & Sustainability 330'
author: 
 - name: Megan Hoover
format: html
---

Load Libraries:
```{r}
# spatial data science
library(tidyverse)
library(sf)
library(units)

# Data
library(AOI)

# Visualization
library(gghighlight)
library(ggrepel)
library(knitr)
```


# 1.1 Define a Projection

For this lab we want to calculate distances between features, therefore we need a projection that preserves distance at the scale of CONUS. For this, we will use the North America Equidistant Conic:
```{r}
eqdc <- '+proj=eqdc +lat_0=40 +lon_0=-96 +lat_1=20 +lat_2=60 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs'

#This PROJ.4 string defines an Equidistant Conic projection with the following parameters:

#+proj=eqdc → Equidistant Conic projection
#+lat_0=40 → Latitude of the projection’s center (40°N)
#+lon_0=-96 → Central meridian (96°W)
#+lat_1=20 → First standard parallel (20°N)
#+lat_2=60 → Second standard parallel (60°N)
#+x_0=0 → False easting (0 meters)
#+y_0=0 → False northing (0 meters)
#+datum=NAD83 → Uses the North American Datum 1983 (NAD83)
#+units=m → Units are in meters
#+no_defs → No additional default parameters from PROJ’s database

```


This projection is commonly used for mapping large areas with an east-west extent, especially in North America, as it balances distortion well between the two standard parallels.

# 1.2 Get USA state boundaries

In R, USA boundaries are stored in the AOI package. In case this package and data are not installed:
```{r}
remotes::install_github("mikejohnson51/AOI")
```
Once installed:

USA state boundaries can be accessed with aoi_get(state = 'conus').
Make sure the data is in a projected coordinate system suitable for distance measurements at the national scale (eqdc).
```{r}
#sf objects with state boundaries
states_conus <- aoi_get(state = "conus")

#check current coordinate system: geographic coordinates in degrees, not suitable for measuring distances.
st_crs(states_conus) #returns WGS 84, coords in degrees

#edqc projection is suitable for distance calculations across CONUS

#Now transform the spatial object to use the EQDC projection:
states_eqdc <- st_transform(states_conus, crs = eqdc)

#check cureent coordinate system now: units should be in m, good for distance calculations
st_crs(states_eqdc)


```


# 1.3 Get country boundaries for Mexico, the United States of America, and Canada

In R, country boundaries are stored in the AOI package.

Country boundaries can be accessed with aoi_get(country = c("MX", "CA", "USA")).
Make sure the data is in a projected coordinate system suitable for distance measurements at the national scale (eqdc).
```{r}
# get country boundaries
country_boundaries<- aoi_get(country = c("MX", "CA", "USA"))

#check structure
print(country_boundaries)

#check coordinate system
st_crs(country_boundaries) # (WGS 84) not ideal for distance.

#transforms your spatial object to the EQDC projection.
countries_eqdc <- st_transform(country_boundaries, crs = eqdc)


```

